---
title: "Final-4"
author: "Mike Lehman"
date: "December 1, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## Introduction

The goal of this assessment is to apply an unsupervised machine learning technique to a given data set to extract new, previously unknown insights from the data. The technique that we will use is a particular approach to clustering. Clustering is simply a way of dividing data into groups of similar items without having been instructed as to what the groups should look like prior to grouping.

Clustering works by plotting all data points into a two-dimensional feature space, then assigning random fixed positions and drawing boundaries which attempt to group all data points in homogenous groups.

There are many different implementations of clustering, the one we will examine is the k-means algorithm. The k-means algorithm uses simple ideas that are easy to understand, but is lacking in sophistication and is driven by a process of random selection that often may require several iterations to extract any meaningful information. The algorithm uses Euclidean distance to determine the spatial distance between two data points to determine which items should be considered homogenous and cluster boundaries drawn.

The algorithm assigns each example to one of k clusters, with k representing a user defined value. The process begins by randomly assigning center points (centroids) during an initial assignment phase, then enters an update phase which involves shifting the cluster centers to a new location in efforts to create more locally homogenous clusters.

## Protein Data Set

The data we will be examining for this assessment come from a breakdown of Eurpopean nations' protein consumption. The data includes a list of 25 different European nations and their respective protein consumption by food group. Each of the good groups has the respective nation's protein consumption in overall percentage. Meaning a value of 10.0 means that 10 percent of that nation's protein consumption comes from that food group.

```{r include=FALSE}
library(fpc)
```

We will begin by reading the data into R. Since we know that one of our features is a list of country names, we will apply stringsAsFactos = FALSE so that the country names read in as character strings instead of factor levels:

```{r}
protein <- read.csv("protein.csv", stringsAsFactors = FALSE)
str(protein)
```

From the output we see that we have a charcter vector for our country labels with all other features being numeric. These refer to the precentage consumption for each food group. Since we are creating distance based clusters using the K-means approach, we will need to scale these features as large value ranges have a significant impact on distance calculations.

First we will create a data frame with only the numeric features:

```{r}
pro.sources <- protein[2:10]
summary(pro.sources)
```

We can see that there are some noticably large value ranges. We can apply the scale function to the entire data frame and store the results in a new data frame to keep our original data intact. R's native scale feature will apply a z-score standardization (how many standard deviations if the value from the mean) to numeric data. 

```{r}
pro.sources_z <- as.data.frame(lapply(pro.sources, scale))
summary(pro.sources_z)
```

From the output we can see that our numeric values are now much closer in range. Now that we have a data frame with standardized numeric features, we can apply K-means clustering to assess relationships amongst the data.

## K-means Clustering

We will make use of the kmeans() function which accepts a data frame of numeric features as well as an additional parameter, the number (k) of clusters centers to consider.

The decision of k values or clusters to consider is arbitrary to a degree. It is often defined based on the domain of the subject matter and multiple values are often tested to achieve the best results. For our assessment we will being with four, considering that the various food groups can be grouped together based on their type: meat, dairy, produce, and other.

Since a great deal of randomization is involved in clustering, we will set a seed value so our results are repeatable.

```{r}
set.seed(123)
protein.clusters <- kmeans(pro.sources_z, 4)
protein.clusters
```

The output of our cluster model shows the frequency of each food source within each cluster. Since the data was scaled, these values refer to how many standard deviations above (positive) or below (negative) the mean. A higher positive value means that that food groups appears frequently within that cluster. 

For example, cluster #1 seems to have many occurrences of Milk and Fish proteins 

Our cluster model has two objects which can be referenced with $ notation to provide more insight into its performance.

```{r}
protein.clusters$size
protein.clusters$centers
```

Size tells us the size of each cluster, meaning how many features appear within that cluster. The cluster centers provide the coordinates of the centroids for each feature.

We can associate cluster assignments back to our original data set as so:

```{r}
protein$cluster <- protein.clusters$cluster
```

Since clustering creates new information, a simple way to assess performance of a clustering model is to relate it back to the original information.

```{r}
protein[, c("cluster","Country")]
```

From this output we can see cluster assignments by country. Previously we saw that cluster 4 had the greatest size, and the output here confirms the same as cluster 4 occurs most frequently. Since this is our largest cluster, it may be helpful to examine some of the characteristics of that cluster.

```{r}
protein[which(protein$cluster == 4), c("cluster","Country")]
```

The countries associated with cluster 4 appear to be widely geographically disbursed, meaning that rules learned about cluster 4 may be broadly applicable to the protein consumption of all of Europe.

Let's return to our data set to examine the different food groups:

```{r}
str(protein)
```

We can run aggregate statistics around each of the food groups by cluster and see what the resulting values tell us about each cluster. For example, we can find the mean percentage of protein consumption for some of the features:

```{r}
aggregate(data = protein, RedMeat ~ cluster, mean)
aggregate(data = protein, Milk ~ cluster, mean)
aggregate(data = protein, Eggs ~ cluster, mean)
```

This first set of aggregate numbers shows us the average percentage of protein consumption by cluster for those foods which we associate with an omnivorous diet. We can also use the same techniques to see numbers for those foods typically associated with vegetarian or non-meat eating diets:

```{r}
aggregate(data = protein, Nuts ~ cluster, mean)
aggregate(data = protein, Fish ~ cluster, mean)
aggregate(data = protein, Starch ~ cluster, mean)
```

It appears that cluster 4 gets more of its protein from read meat and milk whereas clusters 2 and 3 get a noticably higher amount of protein from fish and nuts than cluster 4.

Lastly, it is often useful to plot out the clusters across a two-dimensional feature space to see how cluster assignments are disbursed. Such a plot can illustrate tightly compact groups or those where cluster assignments have significant outliers:

```{r}
plotcluster(protein[2:10], protein$cluster)
```

Clusters 1, 3, and 4 seems to have tightly compact groups, whereas cluster 2 is merely two disparate data points that are somewhat close together but appear to be significantly apart. Recall from earlier that cluster 2 was our smallest so this makes sense.

What this plot also tells us is that the clusters and their respective countries have a tight grouping and not much overalp. Meaning that those countries which belong to cluster 1 have a high degree of homogeneity between each other as far as protein consumption. The same can be said of clusters 3 and 4. Overall protein consumption appears to be highly correlated with a country, meaning that protein consumption can vary greatly depending on which country is observed.

## Conclusion

Using k-means clustering we were able to examine Eurpopean protein consumption to learn some basic rules about the sources of dietary protein for different countries. We learned that there appears to be a high degree of homogeneity within each cluster of countries with very little overlap. This insight would not have been possible by using just the raw data, or at least would have required significant manual searching and sorting.

By applying some basic machine learning concepts we were able to observe previously unknown trends in the data though basic clustering analysis.

[R Source](final-4.R)