---
title: "HW5 - OCR"
author: "Mike Lehman"
date: "October 17, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

A Support Vector Machine is a model that creates an imagined surface with a theoretical boundry between plotted data points that represent feature values. The ultimate goal is to create a flat boundary (hyperplane) that divides the space to create homogenous partitions of data. SVMs can be considered a combination of nearest neighbors classifiers and linear regression models.

The goal of this assessment is to letter characters based on the placement of pixels within an image. To accomplish this task we will apply a Support Vector Machine implementation.

## Data Collection and Preparation

The original data set for this assessment can be found at the [UCI Machine Learning Repository](https://archive.ics.uci.edu/ml/datasets/Letter+Recognition). At this URL there is a data folder link that contains the relevant files.

The file letter-recognition.data contains 20,000 samples with 17 total attributes. 16 of the 17 attributes are numeric values and refer to the spatial position of a box containing a pixel within a larger container that houses the letter. The non-numeric attribute is the target feature, an English letter from A to Z.

The data set we will use for this assessment is nearly identical to the original sample found at the UCI repository.

SVMs require that all features be numeric and that the range be small in scope. All of our input features are already numeric; however, many contain a large range of values. The R package we will be using for the assessment takes care of scaling the data for us, but it is important to note that such a preparatory step is needed.

```{r}
letters <- read.csv("letterdata.csv")
str(letters)
```

The data are already in random order, so we may simply create our train and test data sets by separating based on rows:

```{r}
letters_train <- letters[1:16000, ]
letters_test  <- letters[16001:20000, ]
```

## Train Model

To train our model we will make use of the kernlab R package and its ksvm() function. 

```{r}
library(kernlab)
letter_classifier <- ksvm(letter ~ ., data = letters_train, kernel = "vanilladot")
letter_classifier
```

## Evaluating Model Performance

## Improving Model Performance

## Conclusion



[R Source](hw5-ocr.R) 